A full approximation storage multilevel method \\ for nonlinear variational inequalities

### 2023-06-28 paper/fascd.tex
section 7 fully drafted

This paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive and monotone variational inequalities, with several example problems.  Then Section \ref{sec:cd} reviews the basics of CDs, including the associated iterations, as proposed by Tai \cite{Tai2003}.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, along with special monotone restriction operators \cite{GraeserKornhuber2009}.  Section \ref{sec:cdmultilevel} presents defect constraints \cite{GraeserKornhuber2009}, leading to an apparently new understanding of telescoping sets for up-smoothing, based on an ``incomplete'' CD.  Section \ref{sec:vcycle} then states the FASCD V-cycle, with Section \ref{sec:implementation} addressing implementation concerns, smoother semantics, and F-cycles.  Section \ref{sec:results} gives FASCD numerical results, for problems in Section \ref{sec:vi}, from a parallel Firedrake \cite{Rathgeberetal2016} implementation.  Discussion and outlook in Section \ref{sec:discussion} and an Appendix, which sketchs how FASCD reduces to classical FAS multigrid for PDEs, finish the work.

@@

In this section we consider several practical concerns when implementing and applying \pr{fascd-vcycle} Algorithm \ref{alg:fascd}.  We address convergence criteria and the F-cycle extension of the Algorithm.  We also describe an active set Newton method used in Section \ref{sec:results} as a smoother and coarse solver.

@@

FIXME smoother semantics; shift ``$w^j+y^j$'' etc.

FIXME smoothers use Newton reduced space (RS) \cite{BensonMunson2006}with either direct linear solve or a fixed number of Krylov steps; in latter case either fixed number of CG+ICC for symmetric linear operator (Laplacian), otherwise fixed number of GMRES+ILU

% FIXME try Newton semi-smooth \cite{BensonMunson2006}?

@@

Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} were implemented in Python using the Firedrake library \cite{Rathgeberetal2016}.  In this Section we show numerical results for the several VI Examples in Section \ref{sec:vi}, with a focus on the scaling of FASCD iterations as the mesh resolution increases.  Assuming strong, yet $O(m_J)$ arithmetic, smoothers on each level, as illustrated below, our essential observations are that each V-cycle dramatically reduces the residual norm, and that a couple of F-cycles solves most Examples down to discretization error.

@@

For this experiment we used $P_1$ elements, standard uniform mesh refinement, and a coarsest mesh of $m_0=5^2$ nodes (32 triangles).  The finest mesh had $m_9=2049^2 \approx 4.2 \times 10^{6}$ nodes.  As our smoother we applied a single RS Newton iteration (Section \ref{sec:implementation}), with three (fixed) ICC-preconditioned CG iterations to approximately solve the step equations.  The coarse solver was RS$+$LU, iterated to convergence.  Tolerances \texttt{rtol} $= 10^{-6}$ and \texttt{atol} $= 10^{-12}$ were used in stopping criterion \eqref{eq:stoppingcriterion}.  The initial iterate was set to $w=\max\{0,\psi\}$.

Table \ref{tab:results:classical} shows the resulting iteration counts.  When the RS Newton smoother is applied by itself as a solver it shows rapidly-increasing iterations as expected, roughly proportional to resolution.  By contrast, V-cycle iterations increase slowly, perhaps logarithmically with the resolution.  Noting that for F-cycles we are counting the number of V-cycles after the ramp (Section \ref{sec:implementation}), we see that a constant number of cycles solves the problem.  The spiral problem appears to be only slightly more challenging for these solvers.  The complexity of the coincidence set or free boundary does not strongly-influence FASCD performance.

Each V-cycle does the work of about eight finest-level preconditioned Krylov (CG+ICC) iterations.  For the F-cycle results, where three iterations suffice at all resolutions in either problem, this is close to textbook multigrid efficiency\footnote{Textbook multigrid efficiency is imperfectly defined, but consider \cite{BrownSmithAhmadia2013}: ``an order of magnitude reduction in residual per iteration and solution of the fine-level system at a small multiple of the cost of a residual evaluation.''  However, work of at most ten residual evaluations is often mentioned.} despite the low solution regularity.
\end{example}

@@

Suppose $\Omega=(-3,3)$, $\psi(x) = -0.2|x|$, $\mathcal{K} = \{v \ge \psi\} \subset W^{1,p}(\Omega)$, and $\ell(v) = \int_\Omega g v$ where $g(x)=1$ for $|x|<1$ and $g(x)=-1$ for $|x|>1$.  The exact solution of VI \eqref{eq:vi} is easily computed for any $p$; see Figure \ref{fig:results:plap}.  For this experiment we used $P_1$ elements, a coarsest mesh of $m_0=7$ nodes, and nine standard refinements to a $m_9=3073$ finest mesh.  The default smoother uses three (fixed) iterations of RS Newton, with LU solution of the step equations.\footnote{This is an optimal $O(m_j)$ smoother in 1D.}  The coarse solver is the same but iterated to convergence.  The initial iterate was $w=\psi$, and again \texttt{rtol} $= 10^{-6}$ and \texttt{atol} $= 10^{-12}$.

@@

The $p=1.5$ and $p=4$ cases give different convergence rates in the $\|\cdot\|_\infty$ norm, namely at rates $O(h^{2.005})$ and $O(h^{1.420})$, respectively.
\end{example}

@@

For the experiment we used $P_1$ elements, uniform refinement, and a coarse mesh with $m_0=16^d$ nodes so as to minimally-resolve the coincidence set.  The smoother was two (fixed) RS Newton iterations,  with three preconditioned GMRES iterations to approximately solve the step equations.  The additive Schwarz method (ASM) preconditioner applied incomplete factorization (ILU) preconditioning on the blocks.  The coarse solver was RS Newton and LU, iterated to convergence.  Tolerances \id{rtol} $=10^{-5}$, \id{atol} $=10^{-9}$ were used in \eqref{eq:stoppingcriterion}.

@@

% see pefarrell-fascd/edplay/siaweak.sh|txt
% time in seconds:
%>> 5*60+[11,36,55]
%ans =      311      336      355
\begin{table}[ht]
\begin{tabular}{c@{\hskip 4mm}c@{\hskip 4mm}c@{\hskip 7mm}c@{\hskip 4mm}c}
P & \emph{levels} & $m_J$ & \emph{F-cycles} & \emph{time} (s) \\ \cmidrule{1-5}
 1 & 7 & $1025^2$ & 4 & 311 \\
 4 & 8 & $2049^2$ & 4 & 336 \\
16 & 9 & $4097^2$ & 4 & 355
\end{tabular}
\bigskip
\caption{Iterations and run time (wall clock, in seconds) for parallel FASCD F-cycle solutions of Example \ref{ex:sia}.}
\label{tab:results:siaweak}
\end{table}

@@

\section{Discussion and Conclusion} \label{sec:discussion}

@@

FIXME smoother could be AMG

FIXME desired extension to nonlocal variational inequalities

>>>>

This paper is organized as follows.  In Section \ref{sec:vi} we recall the theory of coercive and monotone variational inequalities, with several example problems.  Section \ref{sec:cd} reviews the basics of CDs, as proposed by Tai \cite{Tai2003}, including the associated iterations.  Section \ref{sec:femultilevel} sets up multilevel FE hierarchies, and then Section \ref{sec:cdmultilevel} presents defect constraints generated via monotone restriction operators \cite{GraeserKornhuber2009}.  Here we propose an apparently new understanding of telescoping sets for up-smoothing, based on an ``incomplete'' CD.  Section \ref{sec:vcycle} then states the FASCD V-cycle, with Section \ref{sec:implementation} addressing convergence criteria, F-cycles, and our smoother choice.  Section \ref{sec:results} gives FASCD numerical results from a parallel Firedrake \cite{Rathgeberetal2016} implementation for the problems introduced in Section \ref{sec:vi}.  Discussion and outlook in Section \ref{sec:discussion}, and an Appendix on how FASCD reduces to classical FAS multigrid for PDEs, finish the work.

@@

In this section we consider several practical concerns when implementing and applying \pr{fascd-vcycle} Algorithm \ref{alg:fascd}.  We address convergence criteria, the F-cycle extension, and an active set Newton method used in Section \ref{sec:results} as a smoother and coarse solver.

@@

\begin{comment}
smoother semantics: in our implementation of the V-cycle we actually shift to solve a VI for $\tilde w^j = w^j+y^j$, then compute $y^j = \tilde w^j - w^j$ after the solve, etc.; this is evidently equivalent
\end{comment}

While a large space of smoothers could be tested,\footnote{See comments in Section \ref{sec:discussion}.} Section \ref{sec:results} only demonstrates one powerful, $O(m_j)$ work per iteration, smoother choice, namely a fixed number of iterations of the active-set Newton method from \cite{BensonMunson2006}, as implemented in PETSc \cite{Balayetal2023}, with a fixed number of preconditioned Krylov iterations to solve the reduced-space step equations.  We will denote this truncated active-set Newton method as \emph{RS Newton}.  The coarse solver is the same, but with LU decomposition solution of the step equations.  For the symmetric Laplacian in Example \ref{ex:results:classical} we used conjugate gradients (CG) with incomplete Cholesky (IC) preconditioning, but for other operators the generalized minimum residual (GMRES) method with incomplete LU (ILU) preconditioning is used.

@@

Algorithms \ref{alg:fascd} and \ref{alg:fascd-fcycle} were implemented in Python using the Firedrake library \cite{Rathgeberetal2016}, with $P_1$ elements and standard, global mesh refinement.  Here we show numerical results for the several VI Examples in Section \ref{sec:vi}, with a focus on the scaling of FASCD iterations as the mesh resolution increases.  Our essential observations are that each V-cycle dramatically reduces the residual norm, and that a few F-cycles solves most Examples down to discretization error.

@@

For this experiment we used a coarsest mesh of $m_0=5^2$ nodes (32 triangles).  The finest mesh had $m_9=2049^2 \approx 4.2 \times 10^{6}$ nodes.  As our smoother we applied a single RS Newton iteration (Section \ref{sec:implementation}), with three (fixed) IC-preconditioned CG iterations to approximately solve the step equations.  The coarse solver was RS$+$LU, iterated to convergence.  Tolerances \texttt{rtol} $= 10^{-6}$ and \texttt{atol} $= 10^{-12}$ were used in stopping criterion \eqref{eq:stoppingcriterion}.  The initial iterate was set to $w=\max\{0,\psi\}$.

Table \ref{tab:results:classical} shows the resulting iteration counts for the various finest meshes of $m_J$ nodes.  When the RS Newton smoother is applied by itself as a solver (``RS only'') it shows rapidly-increasing iterations as expected, roughly proportional to resolution.  By contrast, V-cycle iterations increase slowly, perhaps logarithmically with the resolution.  Noting that for F-cycles we are counting the number of V-cycles after the ramp (Section \ref{sec:implementation}), we see that a constant number of cycles solves the problem.  The spiral problem appears to be only slightly more challenging for these solvers.  The complexity of the coincidence set or free boundary does not strongly-influence FASCD performance.

Each V-cycle does the work of about eight finest-level preconditioned Krylov (CG+IC) iterations.  For the F-cycle results, where three iterations suffice at all resolutions in either problem, this is close to textbook multigrid efficiency\footnote{Textbook multigrid efficiency is imperfectly defined, but consider \cite{BrownSmithAhmadia2013}: ``an order of magnitude reduction in residual per iteration and solution of the fine-level system at a small multiple of the cost of a residual evaluation.''  However, work of at most ten residual evaluations is often mentioned.} despite the low solution regularity.
\end{example}

@@

Suppose $\Omega=(-3,3)$, $\psi(x) = -0.2|x|$, $\mathcal{K} = \{v \ge \psi\} \subset W^{1,p}(\Omega)$, and $\ell(v) = \int_\Omega g v$ where $g(x)=1$ for $|x|<1$ and $g(x)=-1$ for $|x|>1$.  The exact solution of VI \eqref{eq:vi} is easily computed for any $p$; see Figure \ref{fig:results:plap}.  For this experiment we used a coarsest mesh of $m_0=7$ nodes and an $m_9=3073$ finest mesh.  The default smoother uses three (fixed) iterations of RS Newton, with LU solution of the step equations.\footnote{This is an optimal $O(m_j)$ smoother in 1D.}  The coarse solver is the same but iterated to convergence.  The initial iterate was $w=\psi$, and again \texttt{rtol} $= 10^{-6}$ and \texttt{atol} $= 10^{-12}$.

@@

The $p=1.5$ and $p=4$ cases give different convergence rates in the $\|\cdot\|_\infty$ norm, namely at rates $O(h^{2.005})$ and $O(h^{1.420})$, respectively; compare \cite{ChoeLewis1991}.
\end{example}

@@

For the experiment we used a coarse mesh with $m_0=16^d$ nodes so as to minimally-resolve the coincidence set.  The smoother was two (fixed) RS Newton iterations,  with three preconditioned GMRES iterations to approximately solve the step equations.  The additive Schwarz method (ASM) preconditioner applied incomplete factorization (ILU) preconditioning on the blocks.  The coarse solver was RS Newton and LU, iterated to convergence.  Tolerances \id{rtol} $=10^{-5}$, \id{atol} $=10^{-9}$ were used in \eqref{eq:stoppingcriterion}.

@@

% see pefarrell-fascd/edplay/siaweak.sh|txt
% time in seconds:
%>> 5*60+[11,36,55]
%ans =      311      336      355
\begin{table}[ht]
\begin{tabular}{c@{\hskip 4mm}c@{\hskip 4mm}c@{\hskip 7mm}c@{\hskip 4mm}c}
$P$ & \emph{levels} & $m_J$ & \emph{F-cycles} & \emph{time} (s) \\ \cmidrule{1-5}
 1 & 7 & $1025^2$ & 4 & 311 \\
 4 & 8 & $2049^2$ & 4 & 336 \\
16 & 9 & $4097^2$ & 4 & 355
\end{tabular}
\bigskip
\caption{Iterations and run time (wall clock, in seconds) for parallel FASCD F-cycle solutions of Example \ref{ex:sia} on $P$ processes.}
\label{tab:results:siaweak}
\end{table}

@@

\section{Discussion and outlook} \label{sec:discussion}

@@

FIXME smoother could be Newton semi-smooth \cite{BensonMunson2006}; smoother could be interior point method, etc.; smoother could use GMG or AMG on step equations but this is not seen to be necessary in applications so far

FIXME desired extension to nonlocal variational inequalities not tried here
